Variance reduction for Markov chains via martingale representations

### 2020-05-02 aap/Revision/vr-mcmc-revision-new-new.tex
update

In all the sequel, without loss of generality, we assume that $\nabla U(0)=0$ and we set
\begin{equation}
\label{eq:definition-V}
V(x) = 1 + \|x\|^2 \eqsp.
\end{equation}
\begin{lem}
  \label{lem:quadratic_behaviour}
  Assume {\bf (H1)} and {\bf (H2)}. Then there exists $K_2 \geq 0$ such that for any $ \| x \| \geq K_2$ $\ps{\nabla U(x)}{x} \geq (m/2) \| x \|^2$.
\end{lem}
\begin{proof}
  Using {\bf (H1)} and {\bf (H2)}, we have for any $x \in \rset^d$, $\| x \|\geq K_1$,
  \begin{align*}
    \ps{\nabla U(x)}{x}
    &= \int_{0}^{K_1/\| x \|} D^2 U(t x ) [x^{\otimes 2}] \rmd t + \int_{K_1/\| x \|} ^ 1 D^2 U(t x ) [x^{\otimes 2}] \rmd t\\
    & \geq m \|x \|^2 \{1- K_1 (1 +L/m)   / \| x \| \}\eqsp,
  \end{align*}
which proves the first statement. The second statement is obvious.
\end{proof}
\begin{lem}
\label{lem:drift}
 Assume that {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
 Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = m/4$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
 \begin{equation}
\label{eq:lyapunov}
Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma C
\end{equation}
with   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = 25K_2^2/8 + d + m$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $m$ from {\bf (H2)}.
\end{lem}

\begin{proof} 
Note first that
\[
Q_{\gamma}V(x) = 
1+ \| x - \gamma \nabla U(x) \|^2 + \gamma d
\]
Let us first consider the case $x \notin B(0,K_2)$.  
It remains to notice that, by Lemma~\ref{lem:quadratic_behaviour}, we get
\[
\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq \left(1 - \gamma m + 2\gamma^2 L^2\right)\|x\|^2
\]
Since $\gamma < \frac{m}{4L}$, we obtain
$$
Q_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (d + m)
$$
Consider now the case $x \in B(0,K_2)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
$$
Q_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + d + 2(1+L\gamma)^2\|x\|^2\right) \leq
$$

$$
\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(\frac{25K_2^2}{8} + d + m)
$$
\end{proof}

@@

\begin{lem}
\label{lem:v_ergodicity}
 Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(m / 4L^2,1/L)$, for $x \in \rset^d: \|x\| \leq K$  it holds
$$
d_V(\delta_xQ_{\gamma}^{1,n},\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
$$
with $V(x) = 1 + \|x\|^2$ and constants
$$
C = \left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right);
$$

$$
b = \gamma(\frac{25K_1^2}{8} + 2d + m); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + K^2 + 1; \quad \varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right); \quad
$$

$$
\rho = \exp{\left(-\gamma\left(\frac{m}{2} - 2\frac{\frac{25K_1^2}{8} + 2d + m}{K^2 + 1}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right)}
$$
\end{lem}

\begin{proof} Note that the condition {\bf (H1)} implies that for any $x,y \in \rset^d$,
\[
\|x-y-\gamma(\nabla U(x) - \nabla U(y))\|^2 \leq (1 + 2L\gamma + \gamma^2)\|x-y\|^2
\] 
Hence, the result of [Corollary 5]\cite{DurmusBortoli} implies that
\[
\|\delta_xQ^{[1/\gamma]}_{\gamma} - \delta_yQ^{[1/\gamma]}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
\]
with $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$ for all $x,y$ such that $\|x\|,\|y\| \leq K$, which means that the kernel $Q_{\gamma}$ satisfies $([1/\gamma],\varepsilon)$-Doeblin codition on the set $\{x: V(x) < 1+K^2\}$. Together with drift condition \eqref{eq:lyapunov} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.
\end{proof}

>>>>

In all the sequel, without loss of generality, we assume that $\nabla U(0)=0$ and we set
\begin{equation}
\label{eq:definition-V}
V(x) = 1 + \|x\|^2 \eqsp.
\end{equation}
\begin{lem}
  \label{lem:quadratic_behaviour}
  Assume {\bf (H1)} and {\bf (H2)}. Then there exists $K_2 \geq 0$ such that for any $ \| x \| \geq K_2$ $\ps{\nabla U(x)}{x} \geq (m/2) \| x \|^2$.
\end{lem}
\begin{proof}
  Using {\bf (H1)} and {\bf (H2)}, we have for any $x \in \rset^d$, $\| x \|\geq K_1$,
  \begin{align*}
    \ps{\nabla U(x)}{x}
    &= \int_{0}^{K_1/\| x \|} D^2 U(t x ) [x^{\otimes 2}] \rmd t + \int_{K_1/\| x \|} ^ 1 D^2 U(t x ) [x^{\otimes 2}] \rmd t\\
    & \geq m \|x \|^2 \{1- K_1 (1 +L/m)   / \| x \| \}\eqsp,
  \end{align*}
which proves the first statement. The second statement is obvious.
\end{proof}
\begin{lem}
\label{lem:drift}
 Assume {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
 Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = m/2L^2$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
 \begin{equation}
\label{eq:lyapunov}
Q_{\gamma}V(x) \leq \rho^{\gamma} V(x) + \gamma C
\end{equation}
with   $\rho = \exp{\left(-\frac{m}{2}\right)}$, $C = 2K_2^2(L+m) + d + m$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $m$ from {\bf (H2)}.
\end{lem}

\begin{proof} 
Note first that
\[
Q_{\gamma}V(x) = 
1+ \| x - \gamma \nabla U(x) \|^2 + \gamma d
\]
Let us first consider the case $x \notin B(0,K_2)$.  
It remains to notice that, by Lemma~\ref{lem:quadratic_behaviour}, we get
\[
\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq \left(1 - \gamma m + \gamma^2 L^2\right)\|x\|^2
\]
Since $\gamma < \frac{m}{2L^2}$, we obtain
$$
Q_{\gamma}V(x) \leq (1-\gamma m + \gamma^2 L^2)V(x) + \gamma (d + m - \gamma L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (d + m)
$$
Consider now the case $x \in B(0,K_2)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq (1+L\gamma)^2\|x\|^2$, we obtain
$$
Q_{\gamma}V(x) \leq (1-\gamma m + \gamma^2 L^2)V(x) + \gamma\left((m - \gamma L^2)(1+\|x\|^2) + d + L(2+L\gamma)\|x\|^2\right) \leq
$$

$$
\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(2K_2^2(L+m) + d + m)
$$
\end{proof}

@@

\begin{lem}
\label{lem:v_ergodicity}
 Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(m / 4L^2,1/L)$, and for any $x \in \rset^d$  it holds
$$
d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right)
$$
with $V(x) = 1 + \|x\|^2$ and constants
$$
C = \left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right);
$$

$$
b = \gamma(2K_2^2(L+m) + d + m); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + K^2 + 1; \quad \varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right); \quad
$$

$$
\rho = \exp\left\{\left(\frac{m}{2} - 2\frac{2K_2^2(L+m) + d + m}{K^2 + 1}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\}
$$
where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K > 0$ is arbitrary
\end{lem}

\begin{proof} Note that the condition {\bf (H1)} implies that for any $x,y \in \rset^d$,
\[
\|x-y-\gamma(\nabla U(x) - \nabla U(y))\|^2 \leq (1 + 2L\gamma + \gamma^2)\|x-y\|^2
\] 
Hence, the result of [Corollary 5]\cite{DurmusBortoli} implies that
\[
\|\delta_xQ^{[1/\gamma]}_{\gamma} - \delta_yQ^{[1/\gamma]}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
\]
with $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$ for all $x,y$ such that $\|x\|,\|y\| \leq K$, which means that the kernel $Q_{\gamma}$ satisfies $([1/\gamma],\varepsilon)$-Doeblin condition on the set $\{x: V(x) < 1+K^2\}$. Together with drift condition \eqref{eq:lyapunov} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.
\end{proof}
